---
title: "EDA for flights data"
author: "Alexander Kolupaev"
date: "11/04/2017"
output: html_document
---
```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE)

lapply(list("tidyverse","lubridate","magrittr"
,"ggplot2","alr3","gridExtra"), require, c=1)

mymonths <- c("Jan","Feb","Mar", "Apr","May","Jun",
              "Jul","Aug","Sep", "Oct","Nov","Dec")
```

### About
The data set contains information flight delays and performance for domestic carriers in the United States. The year of 2016 is considered in the analysis. The data set can be accessed from [RITA](https://www.transtats.bts.gov/OT_Delay/OT_DelayCause1.asp).

I will only focus on the 4 biggest airlines with combined market share of about [70%](https://www.statista.com/statistics/250577/domestic-market-share-of-leading-us-airlines/). Those are:

* Southwest
* Delta
* American
* United



## Load and Reshape

I am only interested in 4 airlines. Moreover, I only want to look at delays due
to the carrier. Therefore I only leave variables and airlines of interest.
Variables include:

* total flights (arr_flights)
* flights delayed (arr_del15)
* total minutes delayed (arr_delay)
* total minutes delayed because of carrier (carrier_delay)

```{r include=TRUE, echo=FALSE}
df <-
  read.csv("./655176550_12017_3428_airline_delay_causes.csv") %>% ## import data
  filter(grepl("AA|DL|WN|UA", carrier)) %>%                   ## subset
  select(year, arr_del15, arr_flights, X.month, carrier       ## choose columns
       , X.arr_delay, X.carrier_delay, carrier_name, -X) %>%  ##
  na.omit %>%                                                 ## only complete cases
  set_colnames(c("year", "delayed", "arrivals", "month"       ## rename columns
               , "carrier", "delay", "delay_car", "carrier_name")) %>%
  mutate(
    date = ymd(paste(year, month, 1, sep="-")),               ## date column
    carrier = as.factor(as.character(carrier)),               ## factorize
    month = as.integer(month) %>%                             ## factorize
      plyr::mapvalues(1:12, mymonths) %>%
      factor(levels = mymonths)
  )



```

### Summarise
I group by month and carrier and calculate two ratios:

* ratio of flights delayed
* share of delay because of carrier 

```{r include=TRUE, echo=FALSE}

groups <- group_by(df,date, carrier_name)
df1 <- summarise(groups, delayed_flight_share = sum(delayed) / sum(arrivals), carrier_share_min = sum(delay_car) / sum(delay)) %>% ungroup() %>% arrange(date)


```

### Plotting
**we see that July and December have a lot of delays, but its mostly not carriers' fault** - months correspond to peak holiday seasons.
Delta airlines is pretty bad, being responsible for half of delayed time usually.
But somehow they are lucky with weather and other factors, cause proportion of delayed flights is lower than with other carriers.

```{r include=TRUE, echo=FALSE}
p1 <- ggplot(data = df1,
       aes(x = date, y = delayed_flight_share)) +
  geom_line(aes(color = carrier_name)) + scale_x_date(date_breaks = '1 month', date_labels = mymonths)+
  ggtitle ('share of flights delayed')

p2 <- ggplot(data = df1,
       aes(x = date, y = carrier_share_min)) +
  geom_line(aes(color = carrier_name)) + scale_x_date(date_breaks = '1 month', date_labels = mymonths)+
  ggtitle ('share of minutes delayed due to the carrier')

grid.arrange(p1, p2, ncol=1)
```
### Save dataframe
```{r}
write.csv(df1, file="data.csv", row.names=FALSE)
```
